---
title: "Predicting student performance"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(caretEnsemble)
library(corrplot)
library(tidyverse)
```

## Objective

Accurately predict a student's performance (final grade) in math from various attributes (e.g., family size, age, weekly study time, health, etc.) about the student.

## Data

The dataset includes the student's final math score (performance) and 30 different characteristics about the student and is freely available on the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Student+Performance).

```{r load + explore data, message=FALSE, warning=FALSE}
data <- read.csv("~/Documents/Data Science/GitHub_repository/MachineLearningPortfolio/PredictingStudentPerformance/student_data/student-mat.csv", sep = ";", header = TRUE) # student math scores and characteristics

str(data) # 395 students with 33 variables, "G3.y" is the final grade and predictor target

data %>% 
  sapply(function(x) sum(is.na(x))) # no missing data

data %>% 
  boxplot(use.cols = TRUE) # a few fairly skewed predictors and variable scales

```

## Data transformation

After trying out various transformations (including centering, scaling, normalization, BoxCox, and YeoJohnson), I have decided to use standardization which helps to make the data more normal in distribution while still maintaining variation in the predictors.

```{r data transformation, message=FALSE, warning=FALSE}
preprocessParams <- preProcess(data, method = c("center", "scale")) # calculate the pre-process parameters from the dataset by standardizing them
print(preprocessParams) # summarize transform parameters
transformed <- predict(preprocessParams, data) # transform the dataset using the parameters
# summary(transformed) # summarize the transformed dataset
boxplot(transformed, use.cols = TRUE)

```

## Remove redundant features

```{r remove redundant features, message=FALSE, warning=FALSE}
nums <- sapply(transformed, is.numeric)
transformed_numeric <- transformed[ , nums]

factors <- sapply(transformed, is.factor)
transformed_factors <- transformed[ , factors]

correlationMatrix <- cor(transformed_numeric)
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff = 0.75)
print(highlyCorrelated) # remove attributes with an absolute correlation of >= 0.75
corrplot(correlationMatrix)

transformed_pruned <- transformed_numeric %>% 
  select(-c(highlyCorrelated)) 
str(transformed_pruned) # 569 observations with 13 predictors

data <- cbind(transformed_pruned$G3, transformed_pruned[, 1:13], transformed_factors) # do not want G1.y and G2.y (1st and 2nd period grades) since these are highly correlated with the final grade, but make the modeling less useful
 # We just want to predict final grades based off of student attributes and not previous grades
names(data)[1] <- "FinalGrade"

str(data)
```

## Rank features by importance

Based off of importance ranking, the student's number of failures, their mother's and father's education levels, student age, and whether the student wants to pursue higher education are all useful predictors of the student's final grade. Whereas the the student's amount of freetime after school, whether they are involved in extra-curricular activities and their number of school absences all appear to be less helpful predictors.

```{r rank features, message=FALSE, warning=FALSE}
seed <- 23
set.seed(seed)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 3) # prepare training scheme
model <- train(FinalGrade ~ ., data = data, method = "BstLm", preProcess = "scale", trControl = control) # train the model
importance <- varImp(model, scale = FALSE) # estimate variable importance
print(importance) # summarize importance
plot(importance) # plot importance
```

## Feature selection

Feature selection shows that all 30 predictors are needed to minimize the cross-validation root mean squared error. 

```{r feature selection, message=FALSE, warning=FALSE}
set.seed(seed)
control <- rfeControl(functions = rfFuncs, method = "cv", number = 10) # define the control using a random forest selection function
results <- rfe(data[ , -1], data$FinalGrade, sizes = c(1:17), rfeControl = control) # run the RFE algorithm
print(results) # summarize the results
predictors(results) # list the chosen features
selected_predictors <- predictors(results)
plot(results, type=c("g", "o")) # plot the results

#data <- data %>% 
#  select(FinalGrade, selected_predictors)
#str(data)

```

## Spot-checking ML algorithms

Models that appear useful for accurate prediction include gbm, glmnet, glm, svmRadial, rf and C5.0.

{r spot checking algorithms, message=FALSE, warning=FALSE}
control <- trainControl(method = "repeatedcv", number = 10, repeats=3)
set.seed(seed)
metric <- "Accuracy" # define test metric to compare models

algorithms <- c("lda", "glm", "glmnet", "svmRadial", "knn", "nb", "rpart", "C5.0", "treebag", "rf", "gbm")

garbage <- capture.output(models <- caretList(diagnosis ~ ., data = data, trControl = control, methodList = algorithms))
results <- resamples(models)
summary(results)
dotplot(results)
cormodels <- modelCor(results)
print(cormodels)
corrplot(cormodels)


## Tune best ML algorithms

With model tuning, we can slightly increase the prediction accuracies of the best machine learning algorithms. After tuning, the most accurate model is a gradient boosting machine (gbm) with 1,000 trees, a shrinkage parameter of 0.01, and a minimum of 15 observations in the terminal tree nodes, and the model allows for 2-way interactions between predictor variables. This model has an accuracy of 0.9696.  

{r tuning best algorithms, message=FALSE, warning=FALSE}
# glm
modelLookup(model = "glm") # look up the tuning parameter of the model
set.seed(seed)
glm_gridsearch <- train(diagnosis ~ ., data = data, method = "glm", metric = metric, trControl = control)
print(glm_gridsearch)

# gbm
tunegrid <- expand.grid(n.trees = c(900, 1000, 1100), interaction.depth = c(1, 2), 
                        shrinkage = c(0.01), n.minobsinnode = c(15, 20)) 
set.seed(seed)
garbage <- capture.output(gbm_gridsearch <- train(diagnosis ~ ., data = data, method = "gbm", metric = metric,
                                                  tuneGrid = tunegrid, trControl = control))
print(gbm_gridsearch)
max(gbm_gridsearch$results$Accuracy)

# glmnet
tunegrid <- expand.grid(alpha = c(0.5, 0.75, 1), lambda = c(0.0001, 0.0002, 0.0003)) 
set.seed(seed)
glmnet_gridsearch <- train(diagnosis ~ ., data = data, method = "glmnet", metric = metric, tuneGrid = tunegrid,
                       trControl = control)
print(glmnet_gridsearch)
max(glmnet_gridsearch$results$Accuracy)

# svmRadial
tunegrid <- expand.grid(sigma = c(0.14, 0.15, 0.16, 0.17, 0.18), C = c(0.75, 0.80, 0.90, 1)) 
set.seed(seed)
svm_gridsearch <- train(diagnosis ~ ., data = data, method = "svmRadial", metric = metric, tuneGrid = tunegrid,
                       trControl = control)
svm_gridsearch <- train(diagnosis ~ ., data = data, method = "svmRadial", metric = metric, tunelength = 5,
                       trControl = control)
print(svm_gridsearch)
max(svm_gridsearch$results$Accuracy)

# rf
mtry <- sqrt(ncol(data[ , 2:9])) # default for rf tuning parameter is 2.828427
tunegrid <- expand.grid(.mtry = c(1:8))
set.seed(seed)
rf_gridsearch <- train(diagnosis ~ ., data = data, method = "rf", metric = metric, tuneGrid = tunegrid,
                       trControl = control)
print(rf_gridsearch)
max(rf_gridsearch$results$Accuracy)

# C5.0
tunegrid <- expand.grid(.trials = c(5:15), .model = "tree", .winnow = FALSE) 
set.seed(seed)
C5.0_gridsearch <- train(diagnosis ~ ., data = data, method = "C5.0", metric = metric, tuneGrid = tunegrid,
                       trControl = control)
print(C5.0_gridsearch)
max(C5.0_gridsearch$results$Accuracy)


## Ensemble modeling

Stacking the best machine learning models together with a random forest algorithm, slightly improves upon the accuracy of the best tuned gradient boosting machine (which had an accuracy of 0.9696). The accuracy of the stacked model is 0.977, and the model includes the generalized linear model (glm), radial support vector machine (svmRadial), the random forest model, C5.0, and the gradient boosting machine. I did not include the glmnet model since this was highly correlated (> 0.75) with the generalized linear model. 

{r, message=FALSE, warning=FALSE}
algorithmsList2 = c("glm", "svmRadial", "C5.0", "rf", "gbm")

stackControl <- trainControl(method="repeatedcv", number=10, repeats=3, savePredictions=TRUE, classProbs=TRUE)
set.seed(seed)
garbage <- capture.output(models2 <- caretList(diagnosis ~ ., data = data, trControl = stackControl, methodList = algorithmsList2))


set.seed(seed)
stack.glm <- caretStack(models2, method = "glm", metric = "Accuracy", trControl=stackControl) # stack models with glm
  # try out different kinds of models to stack to see which one is most accurate
print(stack.glm) # 0.9726613

set.seed(seed)
stack.rf <- caretStack(models2, method = "rf", metric = "Accuracy", trControl=stackControl) # stack models with glm
  # try out different kinds of models to stack to see which one is most accurate
print(stack.rf) # 0.9769578



To conclude, a stacked machine learning model with eight breast tumor characteristics is quite accurate (0.977) at determining whether the tumor is benign or malignant. This statistical approach could be applied to help improve breast cancer diagnoses and detection.
