---
title: "Predicting student performance"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(caretEnsemble)
library(corrplot)
library(tidyverse)
```

## Objective

Accurately predict a student's performance (final grade) in math from various attributes (e.g., family size, age, weekly study time, health, etc.) about the student.

## Data

The dataset includes the student's final math score (performance) and 30 different characteristics about the student and is freely available on the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Student+Performance).

```{r load + explore data, message=FALSE, warning=FALSE}
studentdata <- read.csv("~/Documents/Data Science/GitHub_repository/MachineLearningPortfolio/PredictingStudentPerformance/student_data/student-mat.csv", sep = ";", header = TRUE) # student math scores and characteristics

str(studentdata) # 395 students with 33 variables, "G3.y" is the final grade and predictor target

studentdata %>% 
  sapply(function(x) sum(is.na(x))) # no missing data

studentdata %>% 
  boxplot(use.cols = TRUE) # a few fairly skewed predictors and variable scales

```

## Data transformation

After trying out various transformations (including centering, scaling, normalization, BoxCox, and YeoJohnson), I have decided to use standardization which helps to make the data more normal in distribution while still maintaining variation in the predictors.

```{r data transformation, message=FALSE, warning=FALSE}
preprocessParams <- preProcess(studentdata, method = c("center", "scale")) # calculate the pre-process parameters from the dataset by standardizing them
print(preprocessParams) # summarize transform parameters
transformed <- predict(preprocessParams, studentdata) # transform the dataset using the parameters
# summary(transformed) # summarize the transformed dataset
boxplot(transformed, use.cols = TRUE)

```

## Remove redundant features

```{r remove redundant features, message=FALSE, warning=FALSE}
nums <- sapply(transformed, is.numeric)
transformed_numeric <- transformed[ , nums]

factors <- sapply(transformed, is.factor)
transformed_factors <- transformed[ , factors]

correlationMatrix <- cor(transformed_numeric)
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff = 0.75)
print(highlyCorrelated) # remove attributes with an absolute correlation of >= 0.75
corrplot(correlationMatrix)

transformed_pruned <- transformed_numeric %>% 
  select(-c(highlyCorrelated)) 
str(transformed_pruned) # 569 observations with 13 predictors

studentdata <- cbind(transformed_pruned$G3, transformed_pruned[, 1:13], transformed_factors) # do not want G1.y and G2.y (1st and 2nd period grades) since these are highly correlated with the final grade, but make the modeling less useful
 # We just want to predict final grades based off of student attributes and not previous grades
names(studentdata)[1] <- "FinalGrade"

str(studentdata)
```

## Rank features by importance

Based off of importance ranking, the student's number of failures, their mother's and father's education levels, student age, and whether the student wants to pursue higher education are all useful predictors of the student's final grade. Whereas the the student's amount of freetime after school, whether they are involved in extra-curricular activities and their number of school absences all appear to be less helpful predictors.

```{r rank features, message=FALSE, warning=FALSE}
seed <- 23
set.seed(seed)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 3) # prepare training scheme
model <- train(FinalGrade ~ ., data = studentdata, method = "BstLm", preProcess = "scale", trControl = control) # train the model
importance <- varImp(model, scale = FALSE) # estimate variable importance
print(importance) # summarize importance
plot(importance) # plot importance
```

## Feature selection

Feature selection shows that all 30 predictors are needed to minimize the cross-validation root mean squared error. 

```{r feature selection, message=FALSE, warning=FALSE}
set.seed(seed)
control <- rfeControl(functions = rfFuncs, method = "cv", number = 10) # define the control using a random forest selection function
results <- rfe(studentdata[ , -1], studentdata$FinalGrade, sizes = c(1:17), rfeControl = control) # run the RFE algorithm
print(results) # summarize the results
predictors(results) # list the chosen features
selected_predictors <- predictors(results)
plot(results, type=c("g", "o")) # plot the results



```

## Spot-checking ML algorithms

Models that appear useful for accurate prediction (minimized root mean squared error) include gbm, rf, treebag, and rpart. In addition these useful models are all fairly correlated with eachother and thus model stacking will likely not improve prediction accuracy.

```{r spot checking algorithms, message=FALSE, warning=FALSE}
control <- trainControl(method = "repeatedcv", number = 10, repeats=3)
set.seed(seed)
metric <- "RMSE" # define test metric to compare models

algorithms <- c("glmboost", "gam", "glmnet", "svmRadial", "mlp", "knn", "rpart", "treebag", "rf", "gbm", "lasso") # regression algorithms

garbage <- capture.output(models <- caretList(FinalGrade ~ ., data = studentdata, trControl = control, methodList = algorithms))
results <- resamples(models)
summary(results)
dotplot(results)
cormodels <- modelCor(results)
print(cormodels)
corrplot(cormodels)
```

## Tune best ML algorithms

With model tuning, we can slightly increase the prediction accuracies of the best machine learning algorithms. After tuning, the most accurate model is a random forest where 14 randomly selected predictor variables are used at each split within the decision tree. This model has a root mean squared error (RMSE) of 0.843.  

```{r tuning best algorithms, message=FALSE, warning=FALSE}
# gbm
modelLookup(model = "gbm")
tunegrid <- expand.grid(n.trees = c(800, 900), interaction.depth = c(1, 2), 
                        shrinkage = c(0.01), n.minobsinnode = c(3, 5)) 
set.seed(seed)

garbage <- capture.output(gbm_gridsearch <- caret::train(FinalGrade ~ ., data = studentdata, method = "gbm", 
                                                  metric = metric, tuneGrid = tunegrid, trControl = control))
print(gbm_gridsearch)
min(gbm_gridsearch$results$RMSE) # 0.8478651


# treebag
modelLookup(model = "treebag")
set.seed(seed)
treebag_mod <- caret::train(FinalGrade ~ ., data = studentdata, method = "treebag", metric = metric,
                            trControl = control)
treebag_gridsearch <- caret::train(FinalGrade ~ ., data = studentdata, method = "treebag", metric = metric, tunelength = 5,
                       trControl = control)
print(treebag_mod) # 0.8593949
print(treebag_gridsearch) # 0.8442427

# rf
modelLookup(model = "rf")
mtry <- sqrt(ncol(studentdata[ , -1])) # default for rf tuning parameter is 5.477226
tunegrid <- expand.grid(.mtry = c(1:15))
set.seed(seed)
rf_gridsearch <- caret::train(FinalGrade ~ ., data = studentdata, method = "rf", metric = metric, tuneGrid = tunegrid,
                       trControl = control)
print(rf_gridsearch)
min(rf_gridsearch$results$RMSE) # 0.8434868

# rpart
modelLookup(model = "rpart")
tunegrid <- expand.grid(.cp = c(0.01, 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 0.1, 0.15)) 
set.seed(seed)
rpart_gridsearch <- caret::train(FinalGrade ~ ., data = studentdata, method = "rpart", metric = metric, tuneGrid = tunegrid,
                       trControl = control)

print(rpart_gridsearch)
min(rpart_gridsearch$results$RMSE) # 0.8702648
```

To conclude, a random forest regression model with all 30 predictor variables (14 randomly selected at each split in the decision tree) is the best model at predicting a student's final test score (performance). While this tuned model surpasses other algorithm types (e.g., generalized linear models, support vector machines) and published studies using the same dataset (Cortez and Silva 2008, best RMSE was 1.75), much of the variation in student performance is still not explained (R squared is 0.296). Thus, likely other factors not explored in this dataset contribute to student performance (e.g., instructor information, socioeconomic variables, class size).

