---
title: "Predicting breast cancer from mass characteristics"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(caretEnsemble)
library(plotly)
source("~/Documents/Data Science/R_Workshop_Hope_041117/panelutils.R")
```

## Objective

Accurately predict whether a breast mass is benign or malignant based off of various characteristics (e.g., symmetry, area, texture, etc.) of the mass.

## Data

The dataset includes the patient diagnosis (212 patients with malignant masses and 357 patients with benign masses) with 30 different mass characteristics. The data were originally from the Univeristy of Wisconsin - Madison and are now freely available on [Kaggle](https://www.kaggle.com/uciml/breast-cancer-wisconsin-data) and the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29).

```{r load + explore data}
data <- read.csv("~/Documents/Data Science/GitHub_repository/MachineLearningPortfolio/PredictingBreastCancer/BreastCancerData.csv")
str(data) # 569 observations with 32 variables
  # an extra column "X" was added to the end of the dataset that needs to be removed

data <- data %>% 
  select(-X)

data %>% 
  sapply(function(x) sum(is.na(x))) # no missing data

data %>% 
  select(-c(id, diagnosis)) %>% 
  boxplot(use.cols = TRUE) # fairly skewed predictors and variable scales

```

## Data transformation

After trying out various transformations (including centering, scaling, normalization, BoxCox, and YeoJohnson), I have decided to use standardization which helps to make the data more normal in distribution while maintaining outliers with large values that could be associated with malignant tumors.

```{r data transformation}
preprocessParams <- preProcess(data[,3:32], method = c("center", "scale")) # calculate the pre-process parameters from the dataset by standardizing them
print(preprocessParams) # summarize transform parameters
transformed <- predict(preprocessParams, data[,3:32]) # transform the dataset using the parameters
# summary(transformed) # summarize the transformed dataset
boxplot(transformed, use.cols = TRUE)

```

## Remove redundant features

```{r remove redundant features}
correlationMatrix <- cor(transformed)
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff = 0.75)
print(highlyCorrelated) # remove attributes with an absolute correlation of >= 0.75

transformed_pruned <- transformed %>% 
  select(-c(highlyCorrelated)) 
str(transformed_pruned) # 569 observations with 12 predictors

data <- cbind(data$diagnosis, transformed_pruned)
names(data)[1] <- "diagnosis"
```

## Rank features by importance

Based off of importance ranking, mean lump area, the standard error of the lump radius, lump texture, and the standard error of the lump concavity are all useful predictors of the tumor diagnosis. Whereas the standard errors of lump symmetry, smoothness, and texture all appear to be less helpful predictors.

```{r rank features}
seed <- 23
set.seed(seed)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 3) # prepare training scheme
model <- train(diagnosis ~ ., data = data, method = "lvq", preProcess = "scale", trControl = control) # train the model
importance <- varImp(model, scale = FALSE) # estimate variable importance
print(importance) # summarize importance
plot(importance) # plot importance
```

## Feature selection

Feature selection shows that eight predictors give the best model accuracy, yet models with just five predictors are similar in accuracy. Thus, if a hospital was interested in streamlining their breast mass measurements, they could probably just assess these five characteristics (area_mean, radius_se, texture_worst, smoothness_worst, and symmytry_worst) and still be able to accurately diagnose the tumor. 

```{r feature selection, message=FALSE, warning=FALSE}
set.seed(seed)
control <- rfeControl(functions = rfFuncs, method = "cv", number = 10) # define the control using a random forest selection function
results <- rfe(data[,2:13], data$diagnosis, sizes = c(1:11), rfeControl = control) # run the RFE algorithm
print(results) # summarize the results
predictors(results) # list the chosen features
selected_predictors <- predictors(results)
plot(results, type=c("g", "o")) # plot the results

data <- data %>% 
  select(diagnosis, selected_predictors)
```

## Spot-checking ML algorithms

Models that appear useful for accurate prediction include gbm, glmnet, glm, svmRadial, rf and C5.0.

```{r spot checking algorithms, message=FALSE, warning=FALSE}
control <- trainControl(method = "repeatedcv", number = 10, repeats=3)
set.seed(seed)
metric <- "Accuracy" # define test metric to compare models

algorithms <- c("lda", "glm", "glmnet", "svmRadial", "knn", "nb", "rpart", "C5.0", "treebag", "rf", "gbm")

garbage <- capture.output(models <- caretList(diagnosis ~ ., data = data, trControl = control, methodList = algorithms))
results <- resamples(models)
summary(results)
dotplot(results)
modelCor(results)
```

## Tune best ML algorithms

```{r}
# glm

# gbm

# glmnet

# svmRadial

# rf

# C5.0
```

## Ensemble modeling

```{r}

```

